On the adhesivity of EGGS

### 2025-02-23 calco/Hypergraphs_with_equivalences.tex
ritocchi

\section{Hypergraphs and term graphs with equivalences}

\begin{definition}
	A \emph{hypergraph with equivalence} $\mathcal{G} = (E_\mathcal{G}, V_{\mathcal{G}}, C_\mathcal{G}, s_\mathcal{G}, t_\mathcal{G}, q_\mathcal{G})$ is a 6-tuple such that $\mathcal{G} = (E_\mathcal{G}, V_{\mathcal{G}}, s_\mathcal{G}, t_\mathcal{G})$ is a hypergraph, $C_\mathcal{G}$ is an object and $q_{\mathcal{G}}: V_{\mathcal{G}}\to C_{\mathcal{G}}$ is a regular epi called \emph{quotient map}. 
	%
	A morphism $h:\mathcal{G\to H}$ is a triple $(h_E, h_V, h_C)$ such that the following diagrams commute
	\[\xymatrix{
		{E_\mathcal{G}}\ar[r]^{s_\mathcal{G}}\ar[d]_{h_E} & {V_{\mathcal{G}}^\star}\ar[d]^{h_V^\star} & {E_\mathcal{G}}\ar[r]^{t_\mathcal{G}}\ar[d]_{h_E} & {V_{\mathcal{G}^\star}}\ar[d]^{h_V^\star} & {V_\mathcal{G}}\ar[r]^{q_\mathcal{G}}\ar[d]_{h_V} & {C_{\mathcal{G}}} \ar[d]^{h_C} \\
		{E_\mathcal{H}}\ar[r]_{s_\mathcal{H}} & {V_{\mathcal{H}}^\star}	& {E_\mathcal{H}}\ar[r]_{t_\mathcal{H}} & {V_{\mathcal{H}^\star}}& {V_\mathcal{H}}\ar[r]_{q_\mathcal{H}} & {C_\mathcal{H}}
	}\]
	The category of hypergraphs with equivalences and their morphisms is denoted $\EqHyp$.

@@

\begin{proof}
	Let $\mathcal{G}$ be a hypergraph, and define $L(\mathcal{G}) := (E_\mathcal{G}, V_{\mathcal{G}}, \{\bullet\}, s_\mathcal{G}, t_\mathcal{G}, !_{V_\mathcal{G}})$. Let now $h: L(\mathcal{G})\to \mathcal{H}$ be a morphism in $\EqHyp$, and consider the following situation.
	\[\xymatrix@C=2.3cm{
		L(\mathcal{G}) \ar@{.>}[d]_{L(f)} \ar[dr]^{h}\\ L(T(\mathcal{H})) \ar[r]_{\epsilon_{\mathcal{H}}} & \mathcal{H}
	}\]
	for $L(T(\mathcal{H}))=(E_\mathcal{H}, V_{\mathcal{H}}, \{\bullet\}, s_\mathcal{H}, t_\mathcal{H}, !_{V_\mathcal{H}})$. Then, $\epsilon_\mathcal{H} = (\id{E_\mathcal{H}}, \id{V_\mathcal{H}}, h_C)$ (by \Cref{rem:eqhyp_morphs}, the last component is uniquely determined by the first two), and $L(f)$ must be $(h_E, h_V, \id{\{\bullet\}})$.
\end{proof}

@@

\begin{corollary}\label{cor:mono1}
	An arrow $h = (h_E, h_V, h_C): \mathcal{G \to H}$ is mono if and only if $T(h)$ is mono.
\end{corollary}

@@

	But the arrow $\initial \to A \times B$ is not epi in general (this is easy to see taking $\Set$ as example), hence such pullback is not preserved by $Q$.
\end{example}

>>>>

\section{Hypergraphs and term graphs with equivalences}
\label{hypereq}

\begin{definition}
	A \emph{hypergraph with equivalence} $\mathcal{G} = (E_\mathcal{G}, V_{\mathcal{G}}, C_\mathcal{G}, s_\mathcal{G}, t_\mathcal{G}, q_\mathcal{G})$ is a 6-tuple such that $\mathcal{G} = (E_\mathcal{G}, V_{\mathcal{G}}, s_\mathcal{G}, t_\mathcal{G})$ is a hypergraph, $C_\mathcal{G}$ is a set and $q_{\mathcal{G}}: V_{\mathcal{G}}\to C_{\mathcal{G}}$ is a regular epi called \emph{quotient map}. 
	%
	A morphism $h:\mathcal{G\to H}$ is a triple $(h_E, h_V, h_C)$ such that the following diagrams commute
	\[\xymatrix{
		{E_\mathcal{G}}\ar[r]^{s_\mathcal{G}}\ar[d]_{h_E} & {V_{\mathcal{G}}^\star}\ar[d]^{h_V^\star} & {E_\mathcal{G}}\ar[r]^{t_\mathcal{G}}\ar[d]_{h_E} & {V_{\mathcal{G}^\star}}\ar[d]^{h_V^\star} & {V_\mathcal{G}}\ar[r]^{q_\mathcal{G}}\ar[d]_{h_V} & {C_{\mathcal{G}}} \ar[d]^{h_C} \\
		{E_\mathcal{H}}\ar[r]_{s_\mathcal{H}} & {V_{\mathcal{H}}^\star}	& {E_\mathcal{H}}\ar[r]_{t_\mathcal{H}} & {V_{\mathcal{H}^\star}}& {V_\mathcal{H}}\ar[r]_{q_\mathcal{H}} & {C_\mathcal{H}}
	}\]
	The category of hypergraphs with equivalences and their morphisms is denoted $\EqHyp$.

@@

\begin{proof}
	Let $\mathcal{G}$ be a hypergraph, and define $L(\mathcal{G}) := (E_\mathcal{G}, V_{\mathcal{G}}, \{\bullet\}, s_\mathcal{G}, t_\mathcal{G}, !_{V_\mathcal{G}})$. Let now $h: L(\mathcal{G})\to \mathcal{H}$ be a morphism in $\EqHyp$, and consider the following situation
	\[\xymatrix@C=2.3cm{
		L(\mathcal{G}) \ar@{.>}[d]_{L(f)} \ar[dr]^{h}\\ L(T(\mathcal{H})) \ar[r]_{\epsilon_{\mathcal{H}}} & \mathcal{H}
	}\]
	for $L(T(\mathcal{H}))=(E_\mathcal{H}, V_{\mathcal{H}}, \{\bullet\}, s_\mathcal{H}, t_\mathcal{H}, !_{V_\mathcal{H}})$. Then, $\epsilon_\mathcal{H} = (\id{E_\mathcal{H}}, \id{V_\mathcal{H}}, h_C)$ (by \Cref{rem:eqhyp_morphs}, the last component is uniquely determined by the first two), and $L(f)$ must be $(h_E, h_V, \id{\{\bullet\}})$.
\end{proof}

@@

\begin{corollary}\label{cor:mono1}
	Let $h = (h_E, h_V, h_C): \mathcal{G \to H}$ be an arrow in  $\EqHyp$. Then it is a mono if and only if $T(h)$ is a mono.
\end{corollary}

@@

	But the arrow $\initial \to A \times B$ is not an epi in general (this is easy to see taking $\Set$ as example), hence such pullback is not preserved by $Q$.
\end{example}

### 2025-02-23 calco/Labeled_hyp_with_eq.tex
ritocchi

As we have done in \Cref{sssect:hyp_alg_sign}, we can define the category of hypergraphs with equivqlence labeled over an algebraic signature.

\begin{definition}
	Let $\Sigma = (O_\Sigma, \ari_\Sigma)$ be an algebraic signature, and let $\mathcal{G}^{\Sigma}$ the hypergraph associated to $\Sigma$.
	Then, the \emph{hypergraph with equivalence} associated to $\Sigma$ is $L(\mathcal{G}^\Sigma)$, and the category of hypergraphs with equivalence labeled over $\Sigma$ is the 
	slice category $\EqHyp_\Sigma := \EqHyp/L(\mathcal{G}^\Sigma)$.
\end{definition}

>>>>

As we have done in \Cref{sssect:hyp_alg_sign}, we can define the category of hypergraphs with equivalence labeled over an algebraic signature.

\begin{definition}
	Let $\Sigma = (O_\Sigma, \ari_\Sigma)$ be an algebraic signature and $\mathcal{G}^{\Sigma}$ the hypergraph associated to $\Sigma$.
	Then the \emph{hypergraph with equivalence} associated to $\Sigma$ is $L(\mathcal{G}^\Sigma)$ and the category of hypergraphs with 
	equivalence labeled over $\Sigma$ is the 
	slice category $\EqHyp_\Sigma = \EqHyp/L(\mathcal{G}^\Sigma)$.
\end{definition}

### 2025-02-23 calco/eggs.tex
ritocchi minimi
ritocchi

\section{EGGs}

@@

\begin{definition}
	Let $\mathcal{G} = (E, V, C, s, t, q)$ be a hypergraph with equivalence, and let $(S, \pi_1, \pi_2)$ be the kernel pair of $q^\star \circ s$.
	Then, $\mathcal{G}$ is an \emph{e-hypergraph} whenever $q^\star \circ t \circ \pi_1 = q^\star \circ t \circ \pi_2$.
	$\egg$ is the full subcategory of $\EqHyp$ where objects are e-hypergraphs, and $I: \egg \to \EqHyp$ is the inclusion functor.
\end{definition}

\begin{lemma}
	$\egg$ has all limits, and $I$ preserves them.
\end{lemma}

\begin{proof}
        Let $D: \cat{I} \to \egg$ be a diagram, with $D(i) = (A_i, B_i, C_i, s_i, t_i, q_i)$, let $(U_i, u_1^i, u_2^i)$ be the kernel pair of $q_i\circ s_i$.
	Let now be $(A, B, C, s, t, q)$, togehter with projections $(\pi_E^i, \pi_V^i, \pi_C^i)_{i \in \cat I}$ the limit of $I \circ D$,
	let $(U, u_1, u_2)$ be the kernel pair of $q\circ s$ and let $(L, (l_i)_{i \in \cat I})$ be the limit of $K \circ I \circ D$.
        By construction (proof of \Cref{prop:eqhyp_complete}), there exists a monomorphism $m: C \to L$ such that $\pi_C^i = l_i \circ m$. Notice that
        \begin{align*}
                q_i^\star \circ s_i\circ \pi^i_E\circ u_1      &= q_i^\star\circ (\pi^i_V)^\star \circ s\circ u_1\\
                                                        &= (\pi_C^i)^\star\circ q^\star\circ s\circ u_1\\
                                                        &=(\pi_C^i)^\star\circ q^\star\circ s\circ u_2\\
                                                        &= q_i^\star \circ s_i \circ \pi_E^i \circ u_2
        \end{align*}
        Then, for each $i$, there exists an arrow $a_i:U\to U_i$ making the following diagram to commute
	\[
                \xymatrix{
			U \ar[r]^{u_1} \ar[d]_{u_2} \ar@{.>}[dr]_{a_i} & A \ar[dr]^{\pi_E^i} & \\
			A \ar[dr]_{\pi_E^i} & U_i \ar[r]^{u_1^i} \ar[d]_{u_2^i}& A_i \ar[d]^{q_i^\star \circ s_i} \\
					     & A_i  \ar[r]_{q_i^\star \circ s_i}& C_i^\star
                }
        \]
        We have then
        \begin{align*}
                l_i^\star\circ m^\star \circ q^\star \circ t \circ u_1    &= q_i^\star\circ (\pi_V^i)^\star \circ t \circ u_1 \\
                                                        &= q_i^\star \circ t_i \circ \pi_E^i \circ u_1 \\
                                                        &= q_i^\star \circ t_i \circ u_1^i \circ a_i \\
                                                        &= q_i^\star \circ t_i \circ u_2^i \circ a_i \\
                                                        &= q_i^\star \circ t_i \circ \pi_E^i \circ u_2 \\
                                                        &= q_i^\star \circ (\pi_V^i)^\star \circ t \circ u_2 \\
                                                        &= l_i^\star \circ m^\star \circ q^\star \circ t \circ u_2
        \end{align*}
        By universal property of limits, we have that \- $m^\star\circ q^\star \circ t \circ u_1 = m^\star \circ q^\star \circ t \circ u_2$, and, since $m$ is mono, $q^\star \circ t \circ u_1 = q^\star \circ t \circ u_2$, hence the thesis.
\end{proof}

@@

\begin{corollary}
	$h: \mathcal{G \to H}$ is a regular mono in $\egg$ if and only if it is a regular mono in $\EqHyp$.
\end{corollary}

>>>>

\section{EGGs}
\label{eggs}

@@

\begin{definition}
	Let $\mathcal{G} = (E, V, C, s, t, q)$ be a hypergraph with equivalence and $(S, \pi_1, \pi_2)$ a kernel pair of $q^\star \circ s$.
	Then, $\mathcal{G}$ is an \emph{e-hypergraph} whenever $q^\star \circ t \circ \pi_1 = q^\star \circ t \circ \pi_2$.
	$\egg$ is the full subcategory of $\EqHyp$ whose objects are e-hypergraphs, and $I: \egg \to \EqHyp$ is the inclusion functor.
\end{definition}

\begin{lemma}
	$\egg$ has all limits and $I$ preserves them.
\end{lemma}

\begin{proof}
        Let $D: \cat{I} \to \egg$ be a diagram, with $D(i) = (A_i, B_i, C_i, s_i, t_i, q_i)$, let $(U_i, u_1^i, u_2^i)$ be the kernel pair of $q_i\circ s_i$.
	Let now be $(A, B, C, s, t, q)$, togehter with projections $(\pi_E^i, \pi_V^i, \pi_C^i)_{i \in \cat I}$ the limit of $I \circ D$,
	let $(U, u_1, u_2)$ be the kernel pair of $q\circ s$ and let $(L, (l_i)_{i \in \cat I})$ be the limit of $K \circ I \circ D$.
        By construction (proof of \Cref{prop:eqhyp_complete}), there exists a mono $m: C \to L$ such that $\pi_C^i = l_i \circ m$. Notice that
        \begin{align*}
                q_i^\star \circ s_i\circ \pi^i_E\circ u_1      &= q_i^\star\circ (\pi^i_V)^\star \circ s\circ u_1\\
                                                        &= (\pi_C^i)^\star\circ q^\star\circ s\circ u_1\\
                                                        &=(\pi_C^i)^\star\circ q^\star\circ s\circ u_2\\
                                                        &= q_i^\star \circ s_i \circ \pi_E^i \circ u_2
        \end{align*}
        Then, for each $i$, there exists an arrow $a_i:U\to U_i$ making the following diagram to commute
	\[
                \xymatrix{
			U \ar[r]^{u_1} \ar[d]_{u_2} \ar@{.>}[dr]_{a_i} & A \ar[dr]^{\pi_E^i} & \\
			A \ar[dr]_{\pi_E^i} & U_i \ar[r]^{u_1^i} \ar[d]_{u_2^i}& A_i \ar[d]^{q_i^\star \circ s_i} \\
					     & A_i  \ar[r]_{q_i^\star \circ s_i}& C_i^\star
                }
        \]
        We have then
        \begin{align*}
                l_i^\star\circ m^\star \circ q^\star \circ t \circ u_1    &= q_i^\star\circ (\pi_V^i)^\star \circ t \circ u_1 \\
                                                        &= q_i^\star \circ t_i \circ \pi_E^i \circ u_1 \\
                                                        &= q_i^\star \circ t_i \circ u_1^i \circ a_i \\
                                                        &= q_i^\star \circ t_i \circ u_2^i \circ a_i \\
                                                        &= q_i^\star \circ t_i \circ \pi_E^i \circ u_2 \\
                                                        &= q_i^\star \circ (\pi_V^i)^\star \circ t \circ u_2 \\
                                                        &= l_i^\star \circ m^\star \circ q^\star \circ t \circ u_2
        \end{align*}
        By universal property of limits, we have that \- $m^\star\circ q^\star \circ t \circ u_1 = m^\star \circ q^\star \circ t \circ u_2$, and, since $m$ is mono, $q^\star \circ t \circ u_1 = q^\star \circ t \circ u_2$, hence the thesis.
\end{proof}

@@

\begin{corollary}
Let $h: \mathcal{G \to H}$ be an arrow in $\egg$. Then it is a regular mono 
%in $\egg$
if and only if $I(h)$ is a regular mono.
% in $\EqHyp$.
\end{corollary}
